---
title: "Math Expressions with Facets in ggplot2"
author: "sahir"
date: "February 8, 2016"
output: html_document
layout: post
tags: [R, ggplot2, facet_wrap, labels]
permalink: facet_wrap_labels
comments: yes
---

```{r setup, echo=FALSE, message=FALSE,warning=FALSE}
library(knitr)
opts_chunk$set(tidy = TRUE, warning = FALSE, message = FALSE)
opts_template$set(
    fig.large = list(fig.width = 7, fig.height = 5, fig.align = 'center', fig.pos = 'h'),
    fig.small = list(fig.width = 6, fig.height = 4, fig.align = 'center', fig.pos = 'h'),
    fig.full = list(fig.width = 9, fig.height = 7, fig.align = 'center', fig.pos = 'h')
)
knit_hooks$set(crop = hook_pdfcrop)
library(plotly)
py <- plotly("sahirbhatnagar", "qpxgmagarx")
```

In this post I show how we can use $$\LaTeX$$ math expressions to label the panels in facets to produce the following plot:

```{r, echo=FALSE, opts.label = 'fig.full'}
library(latex2exp)
data <- devtools::source_gist("https://gist.github.com/sahirbhatnagar/ed3caf50247cae8e3e1c",
                           sha1 = "3073e2ea147ce2818fa4e8841c6efa227378e1aa")$value
appender <- function(string) TeX(paste("$\\log(\\lambda_{\\gamma}) = $",string))  
ggplot(data, aes(log(lambda.beta), ymin = lower, ymax = upper)) + 
  geom_errorbar(color = "grey") + 
  geom_point(aes(x = log(lambda.beta), y = mse), colour = "red") +
  facet_wrap(~lg, scales = "fixed", 
             labeller = as_labeller(appender, default = label_parsed)) + 
  theme(strip.background = element_blank()) + 
    xlab(TeX("$\\log(\\lambda_{\\beta}}$)"))
```




<!--more-->

The updated version of [ggplot2 V 2.0](http://docs.ggplot2.org/dev/index.html) has improved the way we can label panels in [facet plots](http://docs.ggplot2.org/dev/facet_wrap.html) with the use of a [generic labeller](http://docs.ggplot2.org/dev/labeller.html) function. The [latex2exp](https://cran.r-project.org/web/packages/latex2exp/index.html) package has made it much easier to write $$\LaTeX$$ expressions in `R`.

You will need to load the following packages for the code below to work:

1. [devtools](https://cran.r-project.org/web/packages/devtools/index.html)
2. [ggplot2](https://cran.r-project.org/web/packages/ggplot2/)
3. [latex2exp](https://cran.r-project.org/web/packages/latex2exp/index.html)


I have posted some sample data in a [GitHub Gist](https://gist.github.com/sahirbhatnagar/ed3caf50247cae8e3e1c) which you can import into your `R` session using the `source_gist` function from the devtools package:

```{r}
data <- devtools::source_gist("https://gist.github.com/sahirbhatnagar/ed3caf50247cae8e3e1c",
                           sha1 = "3073e2ea147ce2818fa4e8841c6efa227378e1aa")$value
```

Then we create a labelling function which takes as input a string and prepends $$\log(\lambda_{\gamma})$$ to it. Note that `TeX` is the workhorse function that parses $$\LaTeX$$ syntax so that `R` understands it. Otherwise it becomes very messy to try and write more complex math expressions in `R`.

```{r}
appender <- function(string) TeX(paste("$\\log(\\lambda_{\\gamma}) = $",string))  
```

The code to produce the plot above is given by:

```{r, eval=FALSE}
ggplot(data, aes(log(lambda.beta), ymin = lower, ymax = upper)) + 
  geom_errorbar(color = "grey") + 
  geom_point(aes(x = log(lambda.beta), y = mse), colour = "red") +
  facet_wrap(~lg, scales = "fixed", 
             labeller = as_labeller(appender, default = label_parsed)) + 
  theme(strip.background = element_blank()) + 
    xlab(TeX("$\\log(\\lambda_{\\beta}})$"))
```

Note that we need to provide the `default = label_parsed` argument to the `facet_wrap` function so that it interprets the result from the `appender` function as math expressions.




