---
title: "Math Expressions with Facets in ggplot2"
author: "sahir"
date: "February 8, 2016"
output: html_document
layout: post
tags: [R, ggplot2, facet_wrap, labels]
permalink: facet_wrap_labels
comments: yes
---

```{r setup, echo=FALSE, message=FALSE,warning=FALSE}
library(knitr)
opts_chunk$set(tidy = TRUE, warning = FALSE, message = FALSE)
opts_template$set(
    fig.large = list(fig.width = 7, fig.height = 5, fig.align = 'center', fig.pos = 'h'),
    fig.small = list(fig.width = 6, fig.height = 4, fig.align = 'center', fig.pos = 'h'),
    fig.full = list(fig.width = 9, fig.height = 7, fig.align = 'center', fig.pos = 'h')
)
knit_hooks$set(crop = hook_pdfcrop)
library(plotly)
py <- plotly("sahirbhatnagar", "qpxgmagarx")
```

The updated version of [ggplot2 V 2.0](http://docs.ggplot2.org/dev/index.html) has improved the way we can label panels in [facet plots](http://docs.ggplot2.org/dev/facet_wrap.html) with the use of a [generic labeller](http://docs.ggplot2.org/dev/labeller.html) function. 

In this post I show how we can use $\LaTeX$ math expressions to label the panels in facets with the help of the [latex2exp](https://cran.r-project.org/web/packages/latex2exp/index.html) package.

<!--more-->

You will need the following packages for the code below to work:

1. [devtools](https://cran.r-project.org/web/packages/devtools/index.html)
2. [ggplot2](https://cran.r-project.org/web/packages/ggplot2/)
3. [latex2exp](https://cran.r-project.org/web/packages/latex2exp/index.html)


I have posted some sample data in a [GitHub Gist](https://gist.github.com/sahirbhatnagar/ed3caf50247cae8e3e1c) which you can import into your `R` session using the `source_gist` function from the devtools package:

```{r}
data <- devtools::source_gist("https://gist.github.com/sahirbhatnagar/ed3caf50247cae8e3e1c",
                           sha1 = "3073e2ea147ce2818fa4e8841c6efa227378e1aa")$value
```




